Sparse spectral methods for partial differential equations on spherical caps

### 2021-05-05 papers/spherical-caps/revised/spherical-caps.tex
Minor edits and typos, and new figs

Note that we consider partial differential operators that are not necessarily rotational invariant:  for example, one can use these techniques for  Schr\"odinger operators $\DeltaS + v(x,y,z)$ where $v$ is first approximated by a polynomial. A nice feature though is that if the partial differential operator is invariant with respect to rotation around the $z$ axis (e.g., a Schr\"odinger operator with potential $v(z)$) the discretisation decouples, and can be reordered as a block-diagonal matrix. This improves the complexity further to an optimal $O(p^2$), which is demonstrated in Figure \ref{fig:complexity} with $v(x,y,z) = \cos z$.

@@

For the spherical cap, we can use Proposition \ref{prop:construction} to create our one-parameter family of OPs. We first introduce notation for our family of non-classical univariate OPs that will be used as the $r_n$ polynomials above.
\begin{definition}[\cite{snowball2019sparse}]\label{def:OPconstruction}
	Let $\genjacw^{(a,b)}(x)$ be a weight function on the interval $(\alpha, 1)$ given by:
\begin{align*}
	\genjacw^{(a,b)}(x) &:= (x - \alpha)^{a} \: \rho(x)^{b}
\end{align*}
and define the associated inner product by:
\begin{align}
	\ip< p, \: q >_{\genjacw^{(a,b)}} &:= \frac{1}{\normgenjac^{(a,b)}} \: \int_\alpha^1 p(x) \: q(x) \: \genjacw^{(a,b)}(x) \: \D x \label{eqn:ipgenjac}
\end{align}
where
\begin{align}
	\normgenjac^{(a,b)} := \int_\alpha^1 \: \genjacw^{(a,b)}(x) \: \D x \label{eqn:ipnormalisation}
\end{align}
is a normalising constant.
Denote the two-parameter family of orthonormal polynomials on $[\alpha,\beta]$ by $\{\genjac_n^{(a,b)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipgenjac}). That is,
\begin{align}
	\ip< \genjac_n^{(a,b)}, \: \genjac_m^{(a,b)} >_{\genjacw^{(a,b)}} = \delta_{n,m}.
\end{align}
\end{definition}

@@

We can express the three-term recurrences associated with $\genjac_n^{(a,b)}$ as
\begin{align}
	x \genjac_n^{(a,b)}(x) &= \beta_n^{(a,b)} \genjac_{n+1}^{(a,b)}(x) + \alpha_n^{(a,b)} \genjac_n^{(a,b)}(x) + \beta_{n-1}^{(a,b)} \genjac_{n-1}^{(a,b)}(x) \label{eqn:Rrecurrence}
\end{align}
where the coefficients are calculatable (see \cite{snowball2019sparse}). We can use (\ref{eqn:Rrecurrence}) to determine the 3D recurrences for $\scopnkia(x,y,z)$. Importantly, we can deduce sparsity in the recurrence relationships.  We first require the following lemma.

@@

	The 3-term recurrence for multiplication by $z$ follows from (\ref{eqn:Rrecurrence}). For the recurrence for multiplication by $x$, since $\{\scopmjha\}$ for $m = 0,\dots,n+1$, $j = 0,\dots,m$, $h = 0,1$ is an orthogonal basis for any degree $n+1$ polynomial on $\Omega$, we can expand 

@@

A multivariate analogue of Clenshaw's algorithm for recursive evaluation of the OPs has been established for methods on other domains in 2D such as the triangle \cite{olver2019triangle} and the disk-slice \cite{snowball2019sparse}, and we proceed similarly here. Combining each system in (\ref{eqn:jacobimatricesdefinitionalt}) into a block-tridiagonal system, for any $(x,y,z) \in \Omega$, yeilds:
\begin{align*}
	\renewcommand\arraystretch{1.3}
	\begin{pmatrix}
		1 & & & \\
		B_0-G_0(x,y,z) & A_0 & & \\
		C_1 & B_1-G_1(x,y,z) & \quad A_1 \quad & \\
		& C_2 & B_2 - G_2(x,y,z)  & \ddots \\
		& & \ddots &\ddots
	\end{pmatrix}
	\bigscopta(x,y,z) =
		\begin{pmatrix}
	 		\scopa_0 \\ 0 \\ 0 \\ 0 \\ \vdots  \\
		\end{pmatrix},
\end{align*}
where we note $\scopa_0 := \scopa_{0,0,0}(x,y,z) \equiv \genjac_0^{(a,0)} \: \ch_0$, and for each $n = 0,1,2\dots$,
\begin{align*}
	A_n &:= 
		\begin{pmatrix}
			A^{(a)}_{x,n} \\
			A^{(a)}_{y,n} \\
			A^{(a)}_{z,n}
		\end{pmatrix} \in \R^{3(2n+1)\times(2n+3)}, \quad
	C_n := 
		\begin{pmatrix}
			C^{(a)}_{x,n} \\
			C^{(a)}_{y,n} \\
			C^{(a)}_{z,n}
		\end{pmatrix} \in \R^{3(2n+1)\times(2n-1)} \quad (n \ne 0), \nonumber \\
	B_n &:= 
		\begin{pmatrix}
			B^{(a)}_{x,n} \\
			B^{(a)}_{y,n} \\
			B^{(a)}_{z,n}
		\end{pmatrix} \in \R^{3(2n+1)\times(2n+1)}, \quad
	G_n(x,y,z) := 
		\begin{pmatrix}
			xI_{2n+1} \\
			yI_{2n+1} \\
			zI_{2n+1}
		\end{pmatrix} \in \R^{3(2n+1)\times(n+1)}.
\end{align*}

@@

\subsection{Further partial differential operators}\label{subsection:furtherdiffoperators}
General linear partial differential operators with polynomial variable coefficients can be constructed by composing the sparse representations for partial derivatives, conversion between bases, and Jacobi operators. As a canonical example, we can obtain the matrix operator for the $\rho^2$-factored spherical Laplacian $\rho(z)^2 \: \DeltaS$, that will take us from coefficients for expansion in the weighted space $\bigW_N^{(1)}(x,y,z) = \genjacw^{(1,0)}(z) \: \bigscopN^{(1)}(x,y,z)$ to coefficients in the non-weighted space $\bigscopN^{(1)}(x,y,z)$. Note that this construction will ensure the imposition of the Dirichlet zero boundary conditions on $\Omega$, similar to how the Dirichlet zero boundary conditions would be imposed for the operator $\Delta^{(1)}_W$ in Definition \ref{def:differentialoperators}. The matrix operator for this $\rho^2$-factored spherical Laplacian acting on the coefficients vector is then given by
\begin{align*}
	D_\varphi^{(0)} \: W_\varphi^{(1)} + T^{(0)\to(1)} \: T_W^{(1)\to(0)} \: (D_\theta)^2.
\end{align*}
Importantly, this operator will have banded-block-banded structure, and hence will be sparse, as seen in Figure \ref{fig:sparsity}.

Another desirable operator is the Biharmonic operator $\DeltaS^2$, for which we assume zero Dirichlet and Neumann conditions. That is, 
\begin{align*}
	u(x,y,z) &= 0, \quad \pfpx{n}{u}(x,y,z) = \nabla_S u(x,y,z) \cdot \unitvec{n}(x,y,z) = 0 \quad \text{for } (x,y,z) \in \partial\Omega
\end{align*}
where $\partial\Omega$ is the $z=\alpha$ boundary, and $\unitvec{n}(x,y,z)$ is the outward unit normal vector at the point $(x,y,z)$ on the boundary, i.e. $\unitvec{n}(x,y,z) = \unitvec{n}(\vec x) := {\vec x \over \norm{\vec x}} = \vec x$. The matrix operator for the Biharmonic operator will take us from coefficients in the space $\bigW^{(2)}(x,y,z)$ to coefficients in the space $\bigscopN^{(2)}(x,y,z)$. To construct this, we can simply multiply together two of the spherical Laplacian operators defined in \bsrefdef{def:differentialoperators}, namely $\calL^{(0)\to(2)}$ and $\calL_W^{(2)\to(0)}$:
\begin{align*}
	\calB^{(2)}_W := \calL^{(0)\to(2)} \: \calL_W^{(2)\to(0)}.
\end{align*}
Since the operator $\calL_W^{(2)\to(0)}$ acts on coefficients in the $\bigW^{(2)}(x,y,z)$ space, we ensure that we satisfy the zero Dirichlet and Neumann boundary conditions -- such a function could be written $u(x,y,z) = \genjacw^{(2,0)}(z) \: \tilde u(x,y,z)$ and thus its spherical gradient would be zero on the boundary $z = \alpha$. This allows us to apply the $\calL^{(0)\to(2)}$ operator after, safe in the knowledge that boundary conditions have been accounted for. The sparsity and structure of this biharmonic operator are seen in Figure \ref{fig:sparsity}.

@@

\begin{figure}[tp]
	\centering
	\begin{subfigure}[b]{0.55\textwidth}
		\includegraphics[scale=0.4]{condition-numbers-laplacian-N=200}
	\end{subfigure}
	\hfill%

	\begin{subfigure}[b]{0.55\textwidth}
		\includegraphics[scale=0.4]{condition-numbers-max-laplacian}
	\end{subfigure}
	\hfill%
	\caption{Plots of the condition number for the Laplacian operator matrix $\Delta^{(1)}_W$ and the preconditioned matrix $P^{-1} \; \Delta^{(1)}_W$ where $P$ is the matrix of the diagonal of $\Delta^{(1)}_W$. Left: The condition numbers of each diagonal block of $\Delta^{(1)}_W$ for $N = 200$, as well as those for $P^{-1} \; \Delta^{(1)}_W$. Right: The maximum condition number of all diagonal blocks of the Laplacian and the preconditioned Laplacian operators, as the degree $N$ increases.}
	\label{fig:conditionnumbers}
\end{figure}

For a given problem $A\vec{x} = \vec{b}$, the condition number is a measure of how accurate solving the system will be (ignoring any numerical rounding errors), and is given as the condition number of the matrix $A$: 
\begin{align*}
	\kappa(A) := \norm{A}_2 \norm{A^{-1}}_2.
\end{align*}
Recall that the Laplacian operator is block-diagonal. One can therefore look at the condition numbers of each block on the diagonal as proxy for the condition number of the Laplacian (and hence the Poisson equation), where graphical representations are seen in \ref{fig:conditionnumbers} for the $\Delta^{(1)}_W$ operator matrix. As the degree $N$ increases, the condition numbers of these blocks becomes increasingly large (at least, the condition number for the first block where $m=0$ does, which will be the largest for each $N$). Fortunately however, by applying a trivial preconditioning matrix similar to that seen for the ultraspherical method \cite{olver2013fast}, we can significantly bound the condition numbers -- this is also seen in \ref{fig:conditionnumbers}. The preconditioning matrix chosen is $P^{-1}$, where $P$ is the matrix of the diagonal of $\Delta^{(1)}_W$.

@@

We can tackle the problem as follows. Choose an $N \in \N$ large enough for the problem, and denote the coefficient vector for expansion of $u$ in the $\bigWNi$ OP basis up to degree $N$ by $\mathbf{u}$, and the coefficient vector for expansion of $f$ in the $\bigscopNi$ OP basis up to degree $N$ by $\mathbf{f}$. Since $f$ is known, we can obtain $\mathbf{f}$ using the quadrature rule in Section \ref{subsection:expandingfunctions}. In matrix-vector notation, our system hence becomes:
\begin{align*}
    \Delta_W^{(1)} \mathbf{u} = \mathbf{f}
\end{align*}
which can be solved to find $\mathbf{u}$.
In Figure \ref{fig:poisson} we see the solution to the Poisson equation with zero boundary conditions given in (\ref{eqn:poisson}) in the disk-slice $\Omega$. In Figure \ref{fig:poisson} we also show the norms of each block of calculated coefficients of the approximation for four right-hand sides of the Poisson equation with $N = 200$, that is, $(N+1)^2 = 40,401$ unknowns. The right hand sides we choose here are given by
\begin{align*}
	f(x,y,z) &= \norm{\Big(x - (\epsilon + 1 / \sqrt{3}), \: y - (\epsilon + 1 / \sqrt{3}), \: z - (\epsilon + 1 / \sqrt{3})\Big)^\top}
\end{align*}
for differing choices of $\epsilon$ -- this parameter serves to alter the distance from which we would have a singularity. In the plot, a \enquote{block} is simply the group of coefficients corresponding to OPs of the same degree, and so the plot shows how the norms of these blocks decay as the degree of the expansion increases. Thus, the rate of decay in the coefficients is a proxy for the rate of convergence of the computed solution: as typical of spectral methods, we expect the numerical scheme to converge at the same rate as the coefficients decay. We see that we achieve spectral convergence for these examples.

@@

Define $X := (J_x^{(0)})^\top$, $Y := (J_y^{(0)})^\top$, $Z := (J_z^{(0)})^\top$. We can obtain the matrix operator for the variable-coefficient function $v(x,y,z)$ by using the Clenshaw algorithm with matrix inputs as the Jacobi matrices $X, Y, Z$, yielding an operator matrix of the same dimension as the input Jacobi matrices a la the procedure introduced in \cite{olver2019triangle}. We can denote the resulting operator acting on coefficients in the $\bigscopNo$ space by $v(X, Y, Z)$. In matrix-vector notation, our system hence becomes:
\begin{align*}
    (\Delta_W^{(1)} + k^2 \:T^{(0)\to(1)} \: V \: T_W^{(1)\to(0)}) \: \mathbf{u} = \mathbf{f}
\end{align*}
which can be solved to find $\mathbf{u}$. We can see the sparsity and structure of this matrix system in Figure \ref{fig:sparsity} with $v(x,y,z) = zxy^2$ as an example. In Figure \ref{fig:helmholtz} we see the solution to the inhomogeneous variable-coefficient Helmholtz equation with zero boundary conditions given in (\ref{eqn:helmholtz}) in the spherical cap $\Omega$, with $f(x,y,z) = y e^x \genjacw^{(1,0)}(z)$, $v(x,y,z) = 1 - (3(x-x_0)^2 + 5(y-y_0)^2 + 2(z-z_0)^2)$ where $(x_0, z_0) := (0.7, 0.2)$, $y_0 := \sqrt{1 - x_0^2 - z_0^2}$ and $k = 100$. In Figure \ref{fig:helmholtz} we also show the norms of each block of calculated coefficients for the approximation of the solution to the inhomogeneous variable-coefficient Helmholtz equation with various $k$ values. Here, we use $N = 200$, that is, $(N+1)^2 = 40,401$ unknowns. Once again, the rate of decay in the coefficients is a proxy for the rate of convergence of the computed solution, and we see that we achieve spectral convergence.

@@

Our last erxample is the biharmonic equation: find $u(x,y,z)$ given a function $f(x,y,z)$ such that:
\begin{align}
	\begin{cases}
    		\DeltaS^2 u(x,y,z) = f(x,y,z) &\quad \text{in } \Omega \\
		u(x,y,z) = 0, \quad \frac{\partial u}{\partial n}(x,y,z) = \grad_S \: u(x,y,z) \cdot \unitvec{n}(x,y,z) = 0 &\quad \text{on } \partial \Omega
	\end{cases}
	\label{eqn:biharmonic}
\end{align}
where $\DeltaS^2$ is the Biharmonic operator, noting the imposition of zero Dirichlet and Neumann boundary conditions on $u$. For clarity, we reiterate that the unit normal vector in this sense is simply $\unitvec{n}(x,y,z) = \unitvec{n}(\vec x) := {\vec x \over \norm{\vec x}} = \vec x$ (see Section \ref{subsection:furtherdiffoperators}). In Figure \ref{fig:biharmonic} we see the solution to the Biharmonic equation (\ref{eqn:biharmonic}) in the spherical cap $\Omega$. In Figure \ref{fig:biharmonic} we also show the norms of each block of calculated coefficients of the approximation for four more complex right-hand sides of the biharmonic equation with $N = 200$, that is, $(N+1)^2 = 40,401$ unknowns. Once again, the rate of decay in the coefficients is a proxy for the rate of convergence of the computed solution, and we see that we achieve exponential convergence for these more complex functions.

>>>>

Note that we consider partial differential operators that are not necessarily rotational invariant:  for example, one can use these techniques for  Schr\"odinger operators $\DeltaS + v(x,y,z)$ where $v$ is first approximated by a polynomial. A nice feature though is that if the partial differential operator is invariant with respect to rotation around the $z$ axis (e.g., a Schr\"odinger operator with potential $v(z)$) the discretisation decouples, and can be reordered as a block-diagonal matrix. This improves the complexity further to an optimal $O(p^2$), which is demonstrated in Figure~\ref{fig:complexity} with $v(x,y,z) = \cos z$.

@@

For the spherical cap, we can use Proposition \ref{prop:construction} to create our one-parameter family of OPs. We first introduce notation for our family of non-classical univariate OPs that will be used as the $r_n$ polynomials above.
\begin{definition}[\cite{snowball2019sparse}]\label{def:OPconstruction}
	Let $\genjacw^{(a,b)}(x)$ be a weight function on the interval $(\alpha, 1)$ given by:
\begin{align*}
	\genjacw^{(a,b)}(x) &:= (x - \alpha)^{a} \: \rho(x)^{b}
\end{align*}
and define the associated inner product by:
\begin{align}
	\ip< p, \: q >_{\genjacw^{(a,b)}} &:= \frac{1}{\normgenjac^{(a,b)}} \: \int_\alpha^1 p(x) \: q(x) \: \genjacw^{(a,b)}(x) \: \D x \label{eqn:ipgenjac}
\end{align}
where
\begin{align}
	\normgenjac^{(a,b)} := \int_\alpha^1 \: \genjacw^{(a,b)}(x) \: \D x \label{eqn:ipnormalisation}
\end{align}
is a normalising constant.
Denote the two-parameter family of orthonormal polynomials on $[\alpha,\beta]$ by $\{\genjac_n^{(a,b)}\}$, orthonormal with respect to the inner product defined in equation (\ref{eqn:ipgenjac}). That is,
\begin{align}
	\ip< \genjac_n^{(a,b)}, \: \genjac_m^{(a,b)} >_{\genjacw^{(a,b)}} = \delta_{n,m}.
\end{align}
\end{definition}

@@

We can express the three-term recurrences associated with $\genjac_n^{(a,b)}$ as
\begin{align}
	x \genjac_n^{(a,b)}(x) &= \beta_n^{(a,b)} \genjac_{n+1}^{(a,b)}(x) + \alpha_n^{(a,b)} \genjac_n^{(a,b)}(x) + \beta_{n-1}^{(a,b)} \genjac_{n-1}^{(a,b)}(x) \label{eqn:Rrecurrence}
\end{align}
where the coefficients are calculatable (see \cite{snowball2019sparse}). We can use equation (\ref{eqn:Rrecurrence}) to determine the 3D recurrences for $\scopnkia(x,y,z)$. Importantly, we can deduce sparsity in the recurrence relationships.  We first require the following lemma.

@@

	The 3-term recurrence for multiplication by $z$ follows from equation (\ref{eqn:Rrecurrence}). For the recurrence for multiplication by $x$, since $\{\scopmjha\}$ for $m = 0,\dots,n+1$, $j = 0,\dots,m$, $h = 0,1$ is an orthogonal basis for any degree $n+1$ polynomial on $\Omega$, we can expand 

@@

A multivariate analogue of Clenshaw's algorithm for recursive evaluation of the OPs has been established for methods on other domains in 2D such as the triangle \cite{olver2019triangle} and the disk-slice \cite{snowball2019sparse}, and we proceed similarly here. Combining each system in equation (\ref{eqn:jacobimatricesdefinitionalt}) into a block-tridiagonal system, for any $(x,y,z) \in \Omega$, yeilds:
\begin{align*}
	\renewcommand\arraystretch{1.3}
	\begin{pmatrix}
		1 & & & \\
		B_0-G_0(x,y,z) & A_0 & & \\
		C_1 & B_1-G_1(x,y,z) & \quad A_1 \quad & \\
		& C_2 & B_2 - G_2(x,y,z)  & \ddots \\
		& & \ddots &\ddots
	\end{pmatrix}
	\bigscopta(x,y,z) =
		\begin{pmatrix}
	 		\scopa_0 \\ 0 \\ 0 \\ 0 \\ \vdots  \\
		\end{pmatrix},
\end{align*}
where we note $\scopa_0 := \scopa_{0,0,0}(x,y,z) \equiv \genjac_0^{(a,0)} \: \ch_0$, and for each $n = 0,1,2\dots$,
\begin{align*}
	A_n &:= 
		\begin{pmatrix}
			A^{(a)}_{x,n} \\
			A^{(a)}_{y,n} \\
			A^{(a)}_{z,n}
		\end{pmatrix} \in \R^{3(2n+1)\times(2n+3)}, \quad
	C_n := 
		\begin{pmatrix}
			C^{(a)}_{x,n} \\
			C^{(a)}_{y,n} \\
			C^{(a)}_{z,n}
		\end{pmatrix} \in \R^{3(2n+1)\times(2n-1)} \quad (n \ne 0), \nonumber \\
	B_n &:= 
		\begin{pmatrix}
			B^{(a)}_{x,n} \\
			B^{(a)}_{y,n} \\
			B^{(a)}_{z,n}
		\end{pmatrix} \in \R^{3(2n+1)\times(2n+1)}, \quad
	G_n(x,y,z) := 
		\begin{pmatrix}
			xI_{2n+1} \\
			yI_{2n+1} \\
			zI_{2n+1}
		\end{pmatrix} \in \R^{3(2n+1)\times(n+1)}.
\end{align*}

@@

\subsection{Further partial differential operators}\label{subsection:furtherdiffoperators}
General linear partial differential operators with polynomial variable coefficients can be constructed by composing the sparse representations for partial derivatives, conversion between bases, and Jacobi operators. As a canonical example, we can obtain the matrix operator for the $\rho^2$-factored spherical Laplacian $\rho(z)^2 \: \DeltaS$, that will take us from coefficients for expansion in the weighted space $\bigW_N^{(1)}(x,y,z) = \genjacw^{(1,0)}(z) \: \bigscopN^{(1)}(x,y,z)$ to coefficients in the non-weighted space $\bigscopN^{(1)}(x,y,z)$. Note that this construction will ensure the imposition of the Dirichlet zero boundary conditions on $\Omega$, similar to how the Dirichlet zero boundary conditions would be imposed for the operator $\Delta^{(1)}_W$ in Definition \ref{def:differentialoperators}. The matrix operator for this $\rho^2$-factored spherical Laplacian acting on the coefficients vector is then given by
\begin{align*}
	D_\varphi^{(0)} \: W_\varphi^{(1)} + T^{(0)\to(1)} \: T_W^{(1)\to(0)} \: (D_\theta)^2.
\end{align*}
Importantly, this operator will have banded-block-banded structure, and hence will be sparse, as seen in Figure~\ref{fig:sparsity}.

Another desirable operator is the Biharmonic operator $\DeltaS^2$, for which we assume zero Dirichlet and Neumann conditions. That is, 
\begin{align*}
	u(x,y,z) &= 0, \quad \pfpx{n}{u}(x,y,z) = \nabla_S u(x,y,z) \cdot \unitvec{n}(x,y,z) = 0 \quad \text{for } (x,y,z) \in \partial\Omega
\end{align*}
where $\partial\Omega$ is the $z=\alpha$ boundary, and $\unitvec{n}(x,y,z)$ is the outward unit normal vector at the point $(x,y,z)$ on the boundary, i.e. $\unitvec{n}(x,y,z) = \unitvec{n}(\vec x) := {\vec x \over \norm{\vec x}} = \vec x$. The matrix operator for the Biharmonic operator will take us from coefficients in the space $\bigW^{(2)}(x,y,z)$ to coefficients in the space $\bigscopN^{(2)}(x,y,z)$. To construct this, we can simply multiply together two of the spherical Laplacian operators defined in \bsrefdef{def:differentialoperators}, namely $\calL^{(0)\to(2)}$ and $\calL_W^{(2)\to(0)}$:
\begin{align*}
	\calB^{(2)}_W := \calL^{(0)\to(2)} \: \calL_W^{(2)\to(0)}.
\end{align*}
Since the operator $\calL_W^{(2)\to(0)}$ acts on coefficients in the $\bigW^{(2)}(x,y,z)$ space, we ensure that we satisfy the zero Dirichlet and Neumann boundary conditions -- such a function could be written $u(x,y,z) = \genjacw^{(2,0)}(z) \: \tilde u(x,y,z)$ and thus its spherical gradient would be zero on the boundary $z = \alpha$. This allows us to apply the $\calL^{(0)\to(2)}$ operator after, safe in the knowledge that boundary conditions have been accounted for. The sparsity and structure of this biharmonic operator are seen in Figure~\ref{fig:sparsity}.

@@

\begin{figure}[tp]
	\centering
	\begin{subfigure}[b]{0.55\textwidth}
		\includegraphics[scale=0.4]{condition-numbers-laplacian-log-N=200}
	\end{subfigure}
	\hfill%

	\begin{subfigure}[b]{0.55\textwidth}
		\includegraphics[scale=0.4]{condition-numbers-max-laplacian-log}
	\end{subfigure}
	\hfill%
	\caption{Plots of the condition number for the Laplacian operator matrix $\Delta^{(1)}_W$ and the preconditioned matrix $P^{-1} \Delta^{(1)}_W$ where $P$ is the matrix of the diagonal of $\Delta^{(1)}_W$. Top: The condition numbers of each diagonal block of $\Delta^{(1)}_W$ for $N = 200$, as well as those for $P^{-1} \Delta^{(1)}_W$. Bottom: The maximum condition number of all diagonal blocks of the Laplacian and the preconditioned Laplacian operators, as the degree $N$ increases.}
	\label{fig:conditionnumbers}
\end{figure}

For a given problem $A\vec{x} = \vec{b}$, the condition number is a measure of how accurate solving the system will be (ignoring any numerical rounding errors), and is given as the condition number of the matrix $A$: 
\begin{align*}
	\kappa(A) := \norm{A}_2 \norm{A^{-1}}_2.
\end{align*}
Recall that the Laplacian operator is block-diagonal. One can therefore look at the condition numbers of each block on the diagonal as proxy for the condition number of the Laplacian (and hence the Poisson equation), where graphical representations are seen in Figure~\ref{fig:conditionnumbers} for the $\Delta^{(1)}_W$ operator matrix. As the degree $N$ increases, the condition numbers of these blocks becomes grows fast algebraically (at least, the condition number for the first block where $m=0$ does, which will be the largest for each $N$). Fortunately however, by applying a trivial preconditioning matrix similar to that seen for the ultraspherical method \cite{olver2013fast}, we can significantly bound this growth of the condition numbers -- this is also seen in Figure Figure~\ref{fig:conditionnumbers}. The preconditioning matrix chosen is $P^{-1}$, where $P$ is the matrix of the diagonal of $\Delta^{(1)}_W$.

@@

We can tackle the problem as follows. Choose an $N \in \N$ large enough for the problem, and denote the coefficient vector for expansion of $u$ in the $\bigWNi$ OP basis up to degree $N$ by $\mathbf{u}$, and the coefficient vector for expansion of $f$ in the $\bigscopNi$ OP basis up to degree $N$ by $\mathbf{f}$. Since $f$ is known, we can obtain $\mathbf{f}$ using the quadrature rule in Section \ref{subsection:expandingfunctions}. In matrix-vector notation, our system hence becomes:
\begin{align*}
    \Delta_W^{(1)} \mathbf{u} = \mathbf{f}
\end{align*}
which can be solved to find $\mathbf{u}$.
In Figure~\ref{fig:poisson} we see the solution to the Poisson equation with zero boundary conditions given in equation (\ref{eqn:poisson}) in the disk-slice $\Omega$. In Figure~\ref{fig:poisson} we also show the norms of each block of calculated coefficients of the approximation for four right-hand sides of the Poisson equation with $N = 200$, that is, $(N+1)^2 = 40,401$ unknowns. The right hand sides we choose here are given by
\begin{align*}
	f(x,y,z) &= \norm{\Big(x - (\epsilon + 1 / \sqrt{3}), \: y - (\epsilon + 1 / \sqrt{3}), \: z - (\epsilon + 1 / \sqrt{3})\Big)^\top}
\end{align*}
for differing choices of $\epsilon$ -- this parameter serves to alter the distance from which we would have a singularity. In the plot, a \enquote{block} is simply the group of coefficients corresponding to OPs of the same degree, and so the plot shows how the norms of these blocks decay as the degree of the expansion increases. Thus, the rate of decay in the coefficients is a proxy for the rate of convergence of the computed solution: as typical of spectral methods, we expect the numerical scheme to converge at the same rate as the coefficients decay. We see that we achieve spectral convergence for these examples.

@@

Define $X := (J_x^{(0)})^\top$, $Y := (J_y^{(0)})^\top$, $Z := (J_z^{(0)})^\top$. We can obtain the matrix operator for the variable-coefficient function $v(x,y,z)$ by using the Clenshaw algorithm with matrix inputs as the Jacobi matrices $X, Y, Z$, yielding an operator matrix of the same dimension as the input Jacobi matrices a la the procedure introduced in \cite{olver2019triangle}. We can denote the resulting operator acting on coefficients in the $\bigscopNo$ space by $v(X, Y, Z)$. In matrix-vector notation, our system hence becomes:
\begin{align*}
    (\Delta_W^{(1)} + k^2 \:T^{(0)\to(1)} \: V \: T_W^{(1)\to(0)}) \: \mathbf{u} = \mathbf{f}
\end{align*}
which can be solved to find $\mathbf{u}$. We can see the sparsity and structure of this matrix system in Figure~\ref{fig:sparsity} with $v(x,y,z) = zxy^2$ as an example. In Figure~\ref{fig:helmholtz} we see the solution to the inhomogeneous variable-coefficient Helmholtz equation with zero boundary conditions given in equation (\ref{eqn:helmholtz}) in the spherical cap $\Omega$, with $f(x,y,z) = y e^x \genjacw^{(1,0)}(z)$, $v(x,y,z) = 1 - (3(x-x_0)^2 + 5(y-y_0)^2 + 2(z-z_0)^2)$ where $(x_0, z_0) := (0.7, 0.2)$, $y_0 := \sqrt{1 - x_0^2 - z_0^2}$ and $k = 100$. In Figure~\ref{fig:helmholtz} we also show the norms of each block of calculated coefficients for the approximation of the solution to the inhomogeneous variable-coefficient Helmholtz equation with various $k$ values. Here, we use $N = 200$, that is, $(N+1)^2 = 40,401$ unknowns. Once again, the rate of decay in the coefficients is a proxy for the rate of convergence of the computed solution, and we see that we achieve spectral convergence.

@@

Our last erxample is the biharmonic equation: find $u(x,y,z)$ given a function $f(x,y,z)$ such that:
\begin{align}
	\begin{cases}
    		\DeltaS^2 u(x,y,z) = f(x,y,z) &\quad \text{in } \Omega \\
		u(x,y,z) = 0, \quad \frac{\partial u}{\partial n}(x,y,z) = \grad_S \: u(x,y,z) \cdot \unitvec{n}(x,y,z) = 0 &\quad \text{on } \partial \Omega
	\end{cases}
	\label{eqn:biharmonic}
\end{align}
where $\DeltaS^2$ is the Biharmonic operator, noting the imposition of zero Dirichlet and Neumann boundary conditions on $u$. For clarity, we reiterate that the unit normal vector in this sense is simply $\unitvec{n}(x,y,z) = \unitvec{n}(\vec x) := {\vec x \over \norm{\vec x}} = \vec x$ (see Section \ref{subsection:furtherdiffoperators}). In Figure~\ref{fig:biharmonic} we see the solution to the Biharmonic equation (\ref{eqn:biharmonic}) in the spherical cap $\Omega$. In Figure~\ref{fig:biharmonic} we also show the norms of each block of calculated coefficients of the approximation for four more complex right-hand sides of the biharmonic equation with $N = 200$, that is, $(N+1)^2 = 40,401$ unknowns. Once again, the rate of decay in the coefficients is a proxy for the rate of convergence of the computed solution, and we see that we achieve exponential convergence for these more complex functions.
